Life in the canopy: Community trait assessments reveal substantial functional diversity among fern epiphytes

### 2020-01-27 ms/manuscript.Rmd
Revise MS as per submission V2

```{r word-count table}
tibble(
  words = c(
    "Total word count (Introduction, Materials and Methods, Results, Discussion):_6520",
    "Summary:_198",
    "Introduction:_924",
    "Materials and Methods:_2409",
    "Results:_798",
    "Discussion:_2389",
    "Acknowledgements:_170"),
  figs_tabs = c(
    "No. of figures:_7 (all in color)",
    "No. of tables:_5",
    "No. of supporting information files:_8 (Figs. S1-S2; Tables S1-S6)", "", "", "", "")
) %>%
  separate(words, c("words", "word_count"), sep = "_") %>%
  separate(figs_tabs, c("figs", "fig_count"), sep = "_", fill = "right") %>%
  mutate_all(~replace_na(., "")) %>%
  knitr::kable(format = "latex", escape = TRUE, col.names = NULL) %>%
  kable_styling(full_width = FALSE) %>%
  column_spec(1, width = "12em") %>%
  column_spec(2, width = "2em") %>%
  column_spec(3, width = "12em") %>%
  column_spec(4, width = "8em")
```

@@

- The expansion of angiosperm-dominated forests in the Cretaceous and early Cenozoic had a profound effect on terrestrial biota by creating novel ecological niches. The majority of modern fern lineages are hypothesized to have arisen in response to this expansion, particularly fern epiphytes that radiated into the canopy. Recent evidence, however, suggests that epiphytism does not correlate with increased diversification rates in ferns, calling into question the role of the canopy habitat in their diversification.

- To bridge this impasse, we investigated functional traits of fern sporophytes and gametophytes at the species and community level across a broad phylogenetic sampling on the island of Moorea, French Polynesia, including > 120 species and representatives of five major epiphytic radiations.

@@

The appearance of angiosperm-dominated forests in the late Cretaceous and early Cenozoic fundamentally changed climates and had an immense effect on terrestrial ecosystems, including increased global precipitation and the creation of novel niches [@Niklas1983; @Kenrick1997a; @Lloyd2008; @Boyce2010]. 
The heterogeneous niche space of the angiosperm canopy comprising stratified light and humidity levels and a range of substrates for growth became rapidly filled by a variety of epiphytic plants, including mosses [@Fiz-Palacios2011], leafy liverworts [@Feldberg2014], orchids [@Givnish2015b], bromeliads [@Givnish2014a], and ferns [@Schneider2004d]. 
The transition to epiphytic growth was particularly important for modern fern diversification: \circa{ }29% of all fern species are epiphytes vs. \circa{ }10% across all vascular plants [@Kress1986].
Moreover, major transitions to epiphytism occurred at least five times within ferns, each followed by diversification during the late Cretaceous or early Cenozoic [@Schuettpelz2009; @WatkinsJr.2012].
This has led to the hypothesis that epiphytism was a key factor contributing to extant fern diversification [@Schneider2004d; @Schuettpelz2009; @WatkinsJr.2012].
Recent studies exploring this idea [@Sundue2015; @Testo2016a; @Lehtonen2017], however, have largely failed to identify evidence of rapid diversification associated with epiphytism in ferns. 
This raises the possibility that modern fern diversification was not shaped by the evolution of epiphytism *per se*, but rather by the exploitation of various niches within the canopy and on the forest floor.

To better understand the role of the canopy in fern diversification, one obvious direction is to explore putatively adaptive fern epiphyte traits in a phylogenetic context [@Givnish2015]. 
In contrast to most terrestrial plants, epiphytes must contend with greater stress in the form of nutrient limitation, increased evaporative demand, drought, and extreme oscillations in temperature [@Zotz2001; @WatkinsJr.2009; @Lowman2012]. 
Previous studies have identified multiple traits as likely adaptations to these conditions, albeit many without consideration of phylogeny [\eg, @Kluge2007]. 
Epiphytic ferns tend to have morphological features that minimize water loss from their leaves, including shorter stipes [@WatkinsJr.2010], thicker leaves [@Kluge2007; @WatkinsJr.2007b], and protection with scales or hairs [@WatkinsJr.2006b]. 
Additional specialized adaptations to epiphytic growth include humus collection (\eg, *Asplenium*; @Benzing1990, *Drynaria*; @Janssen2005), vegetative desiccation tolerance (\eg, Hymenophyllaceae; @Shreve1911, *Pleopeltis*; @Pessin1925), CAM photosynthesis (\eg, *Pyrrosia*; @Wong1976), and symbiotic relationships with ants (\eg, *Antrophyum*; @WatkinsJr.2008, *Lecanopteris*; @Gay1993a).

While many similar adaptations are found in epiphytic seed plants and serve as key reference points [@Benzing1990], the evolution of epiphytism in ferns is more complex owing to their unique life-cycle. 
Ferns produce a free-living gametophyte that is typically smaller than the sporophyte and relies on aquatic motile sperm.
Unlike the sporophyte, the fern gametophyte lacks vascular tissue, stomata, or cuticle, placing it in constant equilibrium with its external environment. 
In parallel with some of the adaptations in the sporophytes, the gametophytes of some epiphytic ferns also exhibit morphological and physiological adaptations to minimize water loss and tolerate water scarcity. 
For example, extreme desiccation tolerance is prevalent in gametophytes of multiple lineages of epiphytic ferns, but is rare in sporophytes [@WatkinsJr.2007c; @Pittermann2013].
In addition, epiphytic fern gametophytes tend to exhibit more complex three-dimensional morphologies that may function to retain water and slow its loss [@WatkinsJr.2007c].
Despite evidence for adaptations to epiphytism in multiple lineages in both sporophytes and gametophytes, no previous studies to our knowledge have analyzed the functional ecology of fern epiphytes in a broad phylogenetic context including both life stages.

To gain insight into the role of the canopy in promoting or restricting diversity, a functional and community-based approach can be applied to study this question. 
If the canopy is uniformly harsher than the terrestrial environment, thus acting as a 'filter', we would expect functional convergence of epiphytes at the community level, which in turn may result in decreased phylogenetic diversity of epiphytic communities if traits are phylogenetically conserved [@Webb2002; @Cavender-Bares2009]. 
Conversely, the stratification of the canopy may stimulate functional diversification by niche partitioning, leading to increased functional diversity of epiphytic communities. 
Furthermore, the role of the canopy as a promoter or restrictor of diversity may be modulated by overall abiotic conditions, including most notably relative humidity (RH) and temperature. 
For instance, elevation is a strong predictor of fern species richness, with maximal richness occurring at mid-elevations, especially in the tropics [@Kromer2005; @Kessler2011a]. 
Functional diversity is also correlated with species richness [@Schleuter2010]. 
Thus, environmental variation along elevational gradients is expected to impact the functional, phylogenetic, and taxonomic diversity of epiphytic and terrestrial fern communities.

Here, we seek to understand the role of the canopy in fern diversification by investigating traits related to epiphytic growth in fern gametophytes and sporophytes on Moorea, French Polynesia.
Moorea is an isolated tropical oceanic island with high phylogenetic diversity of ferns distributed across a steep elevational gradient, making it ideal for a community phylogenetic analysis of functional traits. 
Specifically, we test the following hypotheses related to epiphytic growth in ferns: 1) Epiphytic ferns have converged upon similar morphologies in both sporophyte and gametophyte phases; 2) Epiphytic fern communities have lower functional and phylogenetic diversity relative to terrestrial communities.

@@

Moorea, French Polynesia (17°32'S, 149°50'W) is a small (\SI{135}{\kilo\meter}^2^) tropical oceanic island more than \SI{5000}{\kilo\meter} from the nearest continental landmass.
This distance acts as a strong barrier to dispersal [@Carson1995; @Dassler2001], which combined with the young age of the island (\circa{ }\SI{1.5}{\ma} [@Duncan1976]) has led to a relatively small, yet phylogenetically diverse fern flora (\circa{ }130 \species, 8/11 orders; taxonomy follows @PteridophytePhylogenyGroupI2016; `r s_table("species-list")`).
Ferns of Moorea include representatives of multiple Cretaceous epiphytic radiations (\eg, *Asplenium*, *Elaphoglossum*, Hymenophylloideae, Polypodiaceae, Vittarioideae; @Schuettpelz2009) and their terrestrial relatives, making them suitable for a comparative study of traits related to the evolution of epiphytic growth in ferns.

@@

A fern community survey was carried out on Moorea comprising 17, \SI{10}{\meter} × \SI{10}{\meter} plots spanning an elevational gradient from \SI{20}{\meter} to \SI{2000}{\meter}, with plots placed \circa every \SI{200}{\meter} [@Nitta2017].
Most plots were located along trails leading to the three main peaks on the island, Mt. Rotui (\SI{899}{\meter}), Mt. Mouaputa (\SI{890}{\meter}), and Mt. Tohiea (\SI{1206}{\meter}).
Each plot was divided into 25, \SI{2}{\meter} × \SI{2}{\meter} subplots.
Presence/absence of all sporophytes in each subplot was scored and summed to produce an abundance ranking for each species per plot from zero (not observed) to 25 (observed in all subplots).
Epiphytic ferns were sampled by hand to \circa{ }\SI{2}{\meter} on tree trunks or confirmed visually from the ground.
Species were categorized as epiphytic or terrestrial based on field observations (see 'Trait selection and measurements' below).
The list of species in each plot was split into epiphytic and terrestrial groups and treated separately during data analysis, hereafter referred to as 'epiphytic' and 'terrestrial' communities.
Vouchers were deposited at GH and UC (abbreviations follow Index Herbariorum; @Thiers2019).

@@

We selected several traits relevant to epiphytic growth (`r table("traits-used")` and `r figure("traits-used-image")`).
We preferred morphological traits over physiological traits because data for morphological traits are easier to gather at the scale of our study (\circa{ }130 \species distributed over an elevational gradient from \SI{200}{\meter} to \SI{2000}{\meter}).
Plant morphology and physiology are closely linked, and physiological traits such as rates of photosynthesis and water use efficiency are known to be correlated with morphological traits such as stipe length in ferns [@WatkinsJr.2010].

@@

Our study reveals a striking duality in the phylogenetic and functional diversity of fern epiphytes. 
On one hand, we find that fern epiphytes are consistently smaller than terrestrial species at both the species and community level (`r table("pics")`, `r figure("cwm-div")`), and that epiphytic communities are phylogenetically clustered (`r figure("comm-div")`).
At the same time, we find that epiphytic fern communities are significantly more diverse functionally compared to terrestrial communities (`r figure("comm-div")`).
Collectively, these results suggest that environmental filtering plays a strong role in limiting plant size, while simultaneously promoting diversification across other trait axes in epiphytic ferns.

@@

Another trait that defied simple classification into epiphytic or terrestrial groups was specific leaf area (SLA; `r figure("cwm-div")`, `r table("pics")`).
Specific leaf area is part of the leaf economic spectrum (LES), which posits that leaf traits including SLA, N and C content, and leaf life span vary along well-defined axes such that plants generally fall into one of two categories: fast-growing species with high SLA, low nutrient content, and short life spans vs. slow-growing species with low SLA, high nutrient content, and long life spans [@Wright2004a; @Diaz2015].
We expected that epiphytes would exhibit lower SLA, given their more extreme growing conditions.
In contrast, we observed overlapping, wide ranging SLA values in both terrestrial and epiphytic communities, with only slightly lower values in epiphytes (`r figure("cwm-div")`).
This may reflect adaptations to specific niches that are not captured in the binary categorization of habitat commonly applied in ferns. 
For example, the following species ranked among those with the five lowest SLA values (\ie, thick, well-protected leaves) on Moorea: *Humata anderssonii* (Mett.) C.Chr. (a twig epiphyte), *Acrostichum aureum* L. (a terrestrial mangrove specialist), and *Paesia divaticatissima* (Dryand.) Copel. (a terrestrial species occurring on high-elevation ridges) (`r s_table("trait-values")`).
Each of these species is likely to experience prolonged periods of drought-stress. 
In-depth characterization of such microhabitats was beyond the scope of our study and should be investigated further in the future.

@@

However, while the canopy clearly acts as a filter, we also find that it simultaneously functions as a promoter of diversity. 
This is evident in our analyses of MPD~func~ and MNTD~func~, which both indicate that epiphytic communities are more functionally diverse compared to terrestrial species (`r figure("comm-div")`), thus contradicting our hypothesis of decreased functional diversity in epiphytic communities.
Additional analyses using different trait combinations support the role of non-size related traits driving this pattern (`r s_figure("corr-plot")`).
We propose that the high functional diversity of epiphytic fern communities is maintained by previously undetected niche partitioning within epiphytic habitats as follows.
The canopy is generally drier than terrestrial habitats, exerting strong selection for small size. 
At the same time, microclimatic conditions in trees vary rapidly over short distances with height and distance from the trunk [@Hietz1995; @Cardelus2005], contributing to greater environmental variation per unit area than terrestrial habitats [@Rojas-Hernandez2018]. 
This vertical and horizontal diversity of niches is occupied by different fern species [@WatkinsJr.2006a; @Cardelus2007; @Parra2009; @Acebey2017].
Thus, when analyzed together, epiphytic fern communities encompass a wider range of morphological variation than terrestrial communities on average.
Assuming that the traits we measured reflect physiological strategies of epiphytes, this high morphological diversity likely derives from physiological adaptations to a variety of within-canopy niches. 
Clearly, fine-grained characterization of canopy niches and associated adaptations of epiphytic ferns occupying them is an important goal for future research.

@@

Recent studies [\eg, @Sundue2015; @Testo2016a; @Lehtonen2017; @Testo2018a] have called into question the scenario of rapid Cretaceous radiations of epiphytic ferns posited by @Schuettpelz2009 by failing to detect increased rates of diversification associated with epiphytic growth.
Our study provides one possible explanation for this finding. 
Here, we demonstrate that epiphytism is likely a more complex trait than previously realized, and almost certainly requires the evolution of multiple associated and dependent traits. 
We propose that the timing and order of the evolution of such traits are related to the complex underlying diversification dynamics observed in these studies. 
Future research investigating the tempo and mode of the evolution of epiphytism in ferns should seek to better partition this diversity when exploring hypotheses of exceptional diversification.

@@

**`r figure("climate")`** Selected microclimatic variables of study plots along an elevational gradient from \SI{200}{\meter} to \SI{1200}{\meter} on Moorea, French Polynesia. Relative humidity (RH) and temperature were recorded every \SI{15}{\minute} with dataloggers from `r final_climate_survey$start` to `r final_climate_survey$end`. Vapor pressure deficit (VPD) was calculated from relative humidity and temperature as described in the methods, and overall means calculated for daily mean, minimum, maximum, and standard deviation (SD). 'Epiphytic' dataloggers were placed at \circa{ }\SI{2}{\meter} on trees, and 'terrestrial' dataloggers placed at ground level. Color indicates growth habit: epiphytic dataloggers in green, terrestrial dataloggers in brown. Trendlines fitted using linear models with an interaction between growth habit and elevation (see Methods); all trendlines significant at \pval{ }< 0.05.

@@

```{r pca, fig.height = 6, fig.width = 6}
pca_plot
ggsave(plot = pca_plot, filename = here::here("results/Fig_3.eps"), height = 6, width = 6, units = "in", device = cairo_ps_high_res)
```

>>>>

```{r word-count table}
tibble(
  words = c(
    "Total word count (Introduction, Materials and Methods, Results, Discussion):_6498",
    "Summary:_199",
    "Introduction:_881",
    "Materials and Methods:_2410",
    "Results:_798",
    "Discussion:_2409",
    "Acknowledgements:_170"),
  figs_tabs = c(
    "No. of figures:_7 (all in color)",
    "No. of tables:_5",
    "No. of supporting information files:_8 (Figs. S1-S2; Tables S1-S6)", "", "", "", "")
) %>%
  separate(words, c("words", "word_count"), sep = "_") %>%
  separate(figs_tabs, c("figs", "fig_count"), sep = "_", fill = "right") %>%
  mutate_all(~replace_na(., "")) %>%
  knitr::kable(format = "latex", escape = TRUE, col.names = NULL) %>%
  kable_styling(full_width = FALSE) %>%
  column_spec(1, width = "12em") %>%
  column_spec(2, width = "2em") %>%
  column_spec(3, width = "12em") %>%
  column_spec(4, width = "8em")
```

@@

- The expansion of angiosperm-dominated forests in the Cretaceous and early Cenozoic had a profound effect on terrestrial biota by creating novel ecological niches. The majority of modern fern lineages are hypothesized to have arisen in response to this expansion, particularly fern epiphytes that radiated into the canopy. Recent evidence, however, suggests that epiphytism does not correlate with increased diversification rates in ferns, calling into question the role of the canopy habitat in fern evolution.

- To understand the role of the canopy in structuring fern community diversity, we investigated functional traits of fern sporophytes and gametophytes across a broad phylogenetic sampling on the island of Moorea, French Polynesia, including > 120 species and representatives of multiple epiphytic radiations.

@@

The appearance of angiosperm-dominated forests in the late Cretaceous and early Cenozoic fundamentally changed climates and had an immense effect on terrestrial ecosystems, including increased global precipitation and the creation of novel niches [@Niklas1983; @Kenrick1997a; @Lloyd2008; @Boyce2010]. 
The heterogeneous niche space of the angiosperm canopy comprising stratified light and humidity levels and a range of substrates for growth became rapidly filled by a variety of epiphytic plants, including mosses [@Fiz-Palacios2011], leafy liverworts [@Feldberg2014], orchids [@Givnish2015b], bromeliads [@Givnish2014a], and ferns [@Schneider2004d]. 
The transition to epiphytic growth was particularly important for modern fern species diversity: \circa{ }29% of all fern species are epiphytes vs. \circa{ }10% across all vascular plants [@Kress1986].
Moreover, major transitions to epiphytism occurred at least five times within ferns, each followed by diversification during the late Cretaceous or early Cenozoic [@Schuettpelz2009; @WatkinsJr.2012].
This has led to the hypothesis that epiphytism was a key factor contributing to extant fern diversification [@Schneider2004d; @Schuettpelz2009; @WatkinsJr.2012].
Recent studies exploring this idea [@Sundue2015; @Testo2016a; @Lehtonen2017; @Testo2018a], however, have largely failed to identify evidence of rapid diversification associated with epiphytism in ferns, calling into question the role of the canopy habitat in fern evolution.

While shifts in diversification rate are relevant, they are only one way epiphytic growth may have influenced fern evolution.
The impact of epiphytic growth could manifest in other ways, such as by shaping phylogenetic and functional diversity at the community level. 
For example, if the canopy is uniformly harsher than the terrestrial environment, we would expect functional convergence of epiphytes at the community level, which in turn may result in decreased phylogenetic diversity of epiphytic communities if traits are phylogenetically conserved [@Webb2002; @Cavender-Bares2009]. 
Conversely, the stratification of the canopy may stimulate functional diversification by niche partitioning, leading to greater functional diversity within epiphytic communities. 
Furthermore, the role of the canopy as a promoter or inhibitor of community diversity may be modulated by overall abiotic conditions, including most notably relative humidity (RH) and temperature. 
For instance, elevation is a strong predictor of fern species richness, with maximal richness occurring at mid-elevations, especially in the tropics [@Kromer2005; @Kessler2011a]. 
Functional diversity is also correlated with species richness [@Schleuter2010]. 
Thus, environmental variation along elevational gradients is expected to impact the functional, phylogenetic, and taxonomic diversity of epiphytic and terrestrial fern communities.

Fern epiphytes possess numerous adaptations that may contribute to their functional diversity.
In contrast to most terrestrial plants, epiphytes must contend with greater stress in the form of nutrient limitation, increased evaporative demand, drought, and extreme oscillations in temperature [@Zotz2001; @WatkinsJr.2009; @Lowman2012].
Epiphytic ferns tend to have morphological features that minimize water loss from their leaves, including shorter stipes [@WatkinsJr.2010], thicker leaves [@Kluge2007; @WatkinsJr.2007b], and protection with scales or hairs [@WatkinsJr.2006b]. 
Additional specialized adaptations to epiphytic growth include humus collection (\eg, *Asplenium*; @Benzing1990, *Drynaria*; @Janssen2005), vegetative desiccation tolerance (\eg, Hymenophyllaceae; @Shreve1911, *Pleopeltis*; @Pessin1925), CAM photosynthesis (\eg, *Pyrrosia*; @Wong1976), and symbiotic relationships with ants (\eg, *Antrophyum*; @WatkinsJr.2008, *Lecanopteris*; @Gay1993a).

While many similar adaptations are found in epiphytic seed plants and serve as key reference points [@Benzing1990], the evolution of epiphytism in ferns is more complex owing to their unique life-cycle. 
Ferns produce a free-living gametophyte that is typically smaller than the sporophyte and relies on aquatic, motile sperm.
Unlike the sporophyte, the fern gametophyte lacks vascular tissue, stomata, or a well developed cuticle, placing it in constant equilibrium with its external environment. 
In parallel with some of the adaptations of sporophytes, the gametophytes of some epiphytic ferns also exhibit morphological and physiological adaptations to minimize water loss and tolerate water scarcity. 
For example, extreme desiccation tolerance is prevalent in gametophytes of multiple lineages of epiphytic ferns, but is rare in sporophytes [@WatkinsJr.2007c; @Pittermann2013].
In addition, epiphytic fern gametophytes tend to exhibit complex three-dimensional morphologies that may function to retain water and slow its loss [@WatkinsJr.2007c].
Despite evidence for adaptations to epiphytism in multiple lineages in both sporophytes and gametophytes, no previous studies to our knowledge have analyzed the functional ecology of fern epiphytes in a broad phylogenetic context including both life stages.

Here, we seek to understand the role of the canopy in structuring fern community diversity by investigating putatively functional traits related to epiphytic growth in fern gametophytes and sporophytes on Moorea, French Polynesia.
Moorea is an isolated tropical oceanic island with high phylogenetic diversity of ferns distributed across a steep elevational gradient, making it ideal for a community phylogenetic analysis of functional traits. 
Specifically, we test the following hypotheses related to epiphytic growth in ferns: 1) Epiphytic ferns have convergent morphologies in both sporophyte and gametophyte phases; 2) Epiphytic fern communities have lower functional and phylogenetic diversity relative to terrestrial communities.

@@

Moorea, French Polynesia (17°32'S, 149°50'W) is a small (\SI{135}{\kilo\meter}^2^) tropical oceanic island more than \SI{5000}{\kilo\meter} from the nearest continental landmass.
This distance acts as a strong barrier to dispersal [@Carson1995; @Dassler2001], which combined with the young age of the island [\circa{ }\SI{1.5}{\ma}; @Duncan1976] has led to a relatively small, yet phylogenetically diverse fern flora [\circa{ }130 \species, 8/11 orders; `r s_table("species-list")`; taxonomy follows @PteridophytePhylogenyGroupI2016].
Ferns of Moorea include representatives of multiple Cretaceous epiphytic radiations [\eg, *Asplenium*, *Elaphoglossum*, Hymenophylloideae, Polypodiaceae, Vittarioideae; @Schuettpelz2009] and their terrestrial relatives, making them suitable for a comparative study of traits related to the evolution of epiphytic growth in ferns.

@@

A fern community survey was carried out on Moorea comprising 17, \SI{10}{\meter} × \SI{10}{\meter} plots spanning an elevational gradient from \SI{20}{\meter} to \SI{2000}{\meter}, with plots placed \circa every \SI{200}{\meter} [@Nitta2017].
Most plots were located along trails leading to the three main peaks on the island, Mt. Rotui (\SI{899}{\meter}), Mt. Mouaputa (\SI{890}{\meter}), and Mt. Tohiea (\SI{1206}{\meter}).
Each plot was divided into 25, \SI{2}{\meter} × \SI{2}{\meter} subplots.
Presence/absence of all sporophytes in each subplot was scored and summed to produce an abundance ranking for each species per plot from zero (not observed) to 25 (observed in all subplots).
Epiphytic ferns were sampled by hand to \circa{ }\SI{2}{\meter} on tree trunks or confirmed visually from the ground.
Species were categorized as epiphytic or terrestrial based on field observations (see 'Trait selection and measurements' below).
The list of species in each plot was split into epiphytic and terrestrial groups and treated separately during data analysis, hereafter referred to as 'epiphytic' and 'terrestrial' communities.
Vouchers were deposited at GH and UC [abbreviations follow Index Herbariorum; @Thiers2019].

@@

We selected several traits putatively relevant to epiphytic growth (`r table("traits-used")` and `r figure("traits-used-image")`).
We preferred morphological traits over physiological traits because data for morphological traits are easier to gather at the scale of our study (\circa{ }130 \species distributed over an elevational gradient from \SI{200}{\meter} to \SI{1200}{\meter}).
Plant morphology and physiology are closely linked, and physiological traits such as rates of photosynthesis and water use efficiency are known to be correlated with morphological traits such as stipe length in ferns [@WatkinsJr.2010].

@@

Our study reveals a striking duality in the phylogenetic and functional diversity of fern epiphytes. 
On one hand, we find that fern epiphytes are consistently smaller than terrestrial species at both the species and community level (`r table("pics")`, `r figure("cwm-div")`), and that epiphytic communities are phylogenetically clustered (`r figure("comm-div")`).
At the same time, we find that epiphytic fern communities are significantly more diverse functionally compared to terrestrial communities (`r figure("comm-div")`).
Collectively, these results suggest that environmental filtering plays a strong role in limiting plant size, while simultaneously supporting functional diversity across other trait axes in epiphytic ferns.

@@

Another trait that defied simple classification into epiphytic or terrestrial groups was specific leaf area (SLA; `r figure("cwm-div")`, `r table("pics")`).
Specific leaf area is part of the leaf economic spectrum (LES), which posits that leaf traits including SLA, N and C content, and leaf life span vary along well-defined axes such that plants generally fall into one of two categories: fast-growing species with high SLA, low nutrient content, and short life spans vs. slow-growing species with low SLA, high nutrient content, and long life spans [@Wright2004a; @Diaz2015].
We expected that epiphytes would exhibit lower SLA, given their more extreme growing conditions.
In contrast, we observed overlapping, wide ranging SLA values in both terrestrial and epiphytic communities, with only slightly lower values in epiphytes when taking elevation into account (`r figure("cwm-div")`).
This may reflect adaptations to specific niches that are not captured by a binary categorization of growth habit.
For example, the following species ranked among those with the five lowest SLA values (\ie, thick, well-protected leaves) on Moorea: *Humata anderssonii* (Mett.) C.Chr. (a twig epiphyte), *Acrostichum aureum* L. (a terrestrial mangrove specialist), and *Paesia divaticatissima* (Dryand.) Copel. (a terrestrial species occurring on high-elevation ridges) (`r s_table("trait-values")`).
Each of these species is likely to experience prolonged periods of drought-stress. 
In-depth characterization of such microhabitats was beyond the scope of our study and should be investigated further in the future.

@@

However, while the canopy clearly acts as a filter, we also find that it simultaneously functions as a promoter of functional diversity. 
This is evident in our analyses of MPD~func~ and MNTD~func~, which both indicate that epiphytic communities are more functionally diverse compared to terrestrial species (`r figure("comm-div")`), thus contradicting our hypothesis of decreased functional diversity in epiphytic communities.
Additional analyses using different trait combinations support the role of non-size related traits driving this pattern (`r s_figure("corr-plot")`).
We propose that the high functional diversity of epiphytic fern communities is maintained by previously undetected niche partitioning within epiphytic habitats as follows.
The canopy is generally drier than terrestrial habitats, exerting strong selection for small size. 
At the same time, microclimatic conditions in trees vary rapidly over short distances with height and distance from the trunk [@Hietz1995; @Cardelus2005], contributing to greater environmental variation per unit area than terrestrial habitats [@Rojas-Hernandez2018]. 
This vertical and horizontal diversity of niches is occupied by different fern species [@WatkinsJr.2006a; @Cardelus2007; @Parra2009; @Acebey2017].
Thus, when analyzed together, epiphytic fern communities encompass a wider range of morphological variation than terrestrial communities on average.
Assuming that the traits we measured reflect physiological strategies of epiphytes, this high morphological diversity likely derives from physiological adaptations to a variety of within-canopy niches. 
Clearly, fine-grained characterization of canopy niches and associated adaptations of epiphytic ferns occupying them is an important goal for future research.

@@

Recent studies [\eg, @Sundue2015; @Testo2016a; @Lehtonen2017; @Testo2018a] have called into question the scenario of rapid Cretaceous radiations of epiphytic ferns posited by @Schuettpelz2009 by failing to detect increased rates of diversification associated with epiphytic growth.
Our study provides a fresh perspective on this issue. 
Here, we demonstrate that epiphytism is likely a complex trait, and almost certainly requires the evolution of multiple associated and dependent traits. 
We propose that the timing and order of the evolution of such traits are related to the conflicting diversification dynamics observed in these studies.
Another possibility raised by our results is that fern epiphytes have radiated morphologically without corresponding species-level diversification [@Givnish2015].
Future research investigating the tempo and mode of the evolution of epiphytism in ferns should seek to better partition this functional diversity when exploring hypotheses of exceptional species diversification.

@@

**`r figure("climate")`** Selected microclimatic variables of study plots along an elevational gradient from \SI{200}{\meter} to \SI{1200}{\meter} on Moorea, French Polynesia. Relative humidity (RH) and temperature were recorded every \SI{15}{\minute} with dataloggers from `r final_climate_survey$start` to `r final_climate_survey$end`. Vapor pressure deficit (VPD) was calculated from relative humidity and temperature as described in the methods, and overall means calculated for daily mean, minimum, maximum, and standard deviation (SD). 'Epiphytic' dataloggers were placed at \circa{ }\SI{2}{\meter} on trees, and 'terrestrial' dataloggers placed at ground level. Color indicates growth habit: epiphytic dataloggers in green, terrestrial dataloggers in brown. Trendlines fitted using linear models with an interaction between growth habit and elevation (see Methods); all trendlines significant at \pval{ }< 0.05. Partially transparent colors indicate outliers excluded from the models (see Methods).

@@

\clearpage
<!-- FIXME: Run the code below from Rstudio within the container, or lines along X-Y axes won't print -->
```{r pca, fig.height = 6, fig.width = 6}
pca_plot
ggsave(plot = pca_plot, filename = here::here("results/Fig_3.eps"), height = 6, width = 6, units = "in", device = cairo_ps_high_res)
```
